Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-12 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from these processes.

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the two hypothetical texts. The figure
indicates that the judge perceives Text B as having significantly higher
quality than Text A. Two key observations support this conclusion: the
positive difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge in this model: it
required estimating more ``unknown'' parameters than the number of
available pairwise comparisons \citep{Thurstone_1927b}. To address this
issue and facilitate the theory's practical applicability, he formulated
five cases derived from the general form, each progressively
incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a ``coarse
scaling'' of traits and to ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from the hypothesis testing process. Specifically, CJ studies
often rely on point estimates of the traits (i.e., the BTL scores)--or
their transformations--to conduct statistical inference. Although
separating these processes simplifies the analysis of CJ data, the
statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
resulting conclusions
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

Thus, to address the limitations of Thurstone's Case V and the BTL
model--particularly their strong assumptions and the disconnect between
trait measurement and hypothesis testing--this study extends Thurstone's
general form using a systematic, integrated approach that combines
Causal and Bayesian inference methods. It begins by formulating a
conceptual model represented by an SCM and a DAG. This model integrates
Thurstone's core theoretical principles--such as the discriminal
processes (i.e., traits) of stimuli--alongside key CJ experimental
design features, including judges' bias, sampling procedures, and
comparison mechanisms. The study then translates the SCM into a bespoke
statistical model that addresses key limitations of the BTL model. In
particular, it enables the analysis of CJ data when the assumptions of
equal dispersion and zero correlation do not hold (see
Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b}),
and when researchers seek to conduct statistical inference or hypothesis
testing (refer to Section~\ref{sec-theory-issue2}). These methodological
innovations potentially enhance the reliability and validity of trait
measurement \citep{Perron_et_al_2015} and improve the accuracy of
statistical conclusions.

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the interactions among all actors and
processes involved in CJ experiments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages create a principled framework for evaluating best practices
in CJ experimental designs--one that better aligns with the demands of
contemporary CJ assessment contexts \citep{Kelly_et_al_2022} and opens
new avenues for research.

For instance,

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical inferences
derived from these processes.

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the two hypothetical texts. The figure
indicates that the judge perceives Text B as having significantly higher
quality than Text A. Two key observations support this conclusion: the
positive difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over A.

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge in this model: it
required estimating more ``unknown'' parameters than the number of
available pairwise comparisons \citep{Thurstone_1927b}. To address this
issue and facilitate the theory's practical applicability, he formulated
five cases derived from this general form, each progressively
incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a ``coarse
scaling'' of traits and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from the hypothesis testing process. Specifically, CJ studies
often rely on point estimates of the traits (i.e., the BTL scores)--or
their transformations--to conduct statistical inference. Although
separating these processes simplifies the analysis of CJ data, the
statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
resulting conclusions
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

Thus, to address the limitations of Thurstone's Case V and the BTL
model--particularly their strong assumptions and the disconnect between
trait measurement and hypothesis testing--this study extends Thurstone's
general form using a systematic, integrated approach that combines
Causal and Bayesian inference methods. The approach begins by
formulating a conceptual model represented by an SCM and a DAG. This
model integrates Thurstone's core theoretical principles--such as the
discriminal processes (i.e., traits) of stimuli--alongside key CJ
experimental design features, including judges' bias, sampling
procedures, and comparison mechanisms.

The study then translates the SCM into a bespoke statistical model that
addresses key limitations of the BTL model. This model allows
researchers to analyze CJ data when the assumptions of equal dispersion
and zero correlation do not hold (see Section~\ref{sec-theory-issue1a}
and Section~\ref{sec-theory-issue1b}), and when they need to perform
statistical inference (refer to Section~\ref{sec-theory-issue2}). In
particular, the model accounts for judge biases, captures the
hierarchical structure of stimuli, accommodates heterogeneity in
discriminal dispersions, and incorporates measurement error in the
hypothesis testing process. These methodological innovations can
potentially enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
inferences.

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the interactions among all actors and
processes involved in CJ experiments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages create a principled framework for evaluating best practices
in CJ experimental designs, one that better aligns with the demands of
contemporary CJ assessment contexts \citep{Kelly_et_al_2022} and opens
new avenues for research.

Five research avenues deserve particular attention due to their direct
impact on the reliability and validity of trait estimates and the
accuracy of statistical inferences. First, researchers should
investigate how sampling and comparison mechanisms influence CJ
experimental outcomes and affect the robustness of statistical
inferences. Second, further work is needed to understand how judges'
biases impact the number of comparisons required to achieve reliable
trait estimates. Third, researchers should re-evaluate existing methods
for identifying and excluding `misfitting' judges and stimuli to ensure
their effectiveness. Fourth, studies should explore how the hierarchical
structure of stimuli influences both experimental design and hypothesis
testing in CJ contexts. Finally, studies should explore principled
approaches for selecting contextually relevant variables.

{About the implications of the sampling and comparison mechanisms}

{About judges biases, judgment discrepancies, and the number of judges
and judgments}

{About removing misfit judges and stimuli}

{About the hierarchical structure of stimuli}

{About revisit the approaches for selecting contextually relevant
variables}
